---
title: "Build a Mini Reference Library"
author: "Win Cowger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a Mini Reference Library}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)

data.table::setDTthreads(2)
```

```{r setup}
library(OpenSpecy)
```

This example combines a few files bundled with OpenSpecy into a small reference
library. Real library builds usually need larger lookup tables and more
curation, but the same helper functions apply.

## Read And Combine Spectra

`c_spec()` and `build_lib()` default to the widest range represented by the
source spectra at resolution 6. Values outside each source's original range are
kept as `NA`, so useful spectral regions are not discarded.
This compact vignette example uses the shared overlapping range to keep the
rendered output small.

```{r mini-library-read}
mini_files <- c(
  read_extdata("raman_hdpe.csv"),
  read_extdata("ftir_ldpe_soil.asp"),
  read_extdata("raman_atacamit.spc")
)

mini_sources <- lapply(mini_files, read_any)
mini_sources <- lapply(mini_sources, function(x) {
  x$metadata$intensity_units <- "absorbance"
  attr(x, "intensity_unit") <- "absorbance"
  x
})
mini_raw <- c_spec(mini_sources, range = "common", res = 6)

check_OpenSpecy(mini_raw)
dim(mini_raw$spectra)
mini_raw$metadata[, "file_name", with = FALSE]
```

## Create And Fill A Lookup

Lookup templates help users see which metadata values need curation. When
`path` is not supplied, the template is returned as a `data.table`; users can
write it to CSV by supplying `path`. Lookup joins use exact values, so edit the
template until the key column matches the object metadata.

```{r mini-library-template}
template <- make_lib_lookup_template(
  mini_raw,
  columns = "file_name",
  add = c("material", "material_type")
)

template
```

For this small example, create the lookup table directly in R. The same shape
could come from a CSV edited outside R.

```{r mini-library-join}
lookup <- data.table::data.table(
  file_name = basename(mini_files),
  library_type = c("example", "example", "example"),
  spectrum_type = c("raman", "ftir", "raman"),
  material = c("hdpe", "ldpe in soil", "atacamite")
)

hierarchy <- data.table::data.table(
  material = c("hdpe", "ldpe in soil", "atacamite"),
  material_class = c("polyethylene", "polyethylene", "copper mineral"),
  material_type = c("plastic", "plastic", "mineral")
)

join_lib_metadata(mini_raw, lookup, by = "file_name",
                  require_complete = TRUE)$metadata[
  , c("file_name", "library_type", "spectrum_type", "material"),
  with = FALSE
]
```

## Build A Mini Library

`build_lib()` can run the ordinary lookup and material hierarchy joins before
applying named recipes. Recipe names become names in the returned list. Empty
recipes keep the merged spectra unchanged; other recipe lists are passed to
`process_spec()`. Missing values and processing attributes are handled
automatically. Metadata column names are also cleaned to lowercase underscore
names. Known aliases are coalesced using an editable lookup table. Variants
that differ only by underscores or one terminal plural `s` match
automatically.

Before sources are merged, `build_lib()` also converts declared reflectance and
transmittance spectra to absorbance by default. A nonempty
`attr(x, "intensity_unit")` is the primary truth for the whole object;
otherwise, `metadata$intensity_units` is evaluated spectrum by spectrum.
Unknown or missing units are left unchanged with a warning. Use
`convert_intensity = FALSE` when unit handling has already been completed
outside the builder.

The normal input is one or more file paths; one `OpenSpecy` or a list of
`OpenSpecy` objects supports sources already loaded in memory. An RDS path may
store either one object or a list of objects. Other formats are read with
`read_any()`. Large same-axis source lists are prepared in bulk, which is much
faster for legacy RDS files that store many one-spectrum objects. Named stages
and elapsed time are reported while the library is built; use
`progress = FALSE` when quiet output is preferred. Supplying
`restrict_range_args` triggers the existing `restrict_range()` operation before
deduplication and recipes; multiple retained ranges can exclude a known silent
region without custom workflow code.

```{r metadata-name-lookup}
name_lookup <- lib_metadata_name_lookup(
  project_code = c("campaign id", "study code"),
  regex = list(instrument_mode = "^method_[0-9]+$")
)
name_lookup[
  canonical_name %in% c("material_color", "number_of_accumulations")
]
lib_clean_name(c("User Name", "Laser (%)", "Method...3"))
```

Named arguments add exact aliases to the defaults, while `regex` adds patterns
evaluated against cleaned names. Overlapping regex patterns produce an error
that identifies the source column and matching rules. Pass the result as
`metadata_name_lookup`. Set `clean_metadata_values = TRUE` in `build_lib()`
or `clean_values = TRUE` in `lib_clean_metadata()` to lowercase, trim, and
ASCII-normalize character metadata values before joins. Ordinary and
hierarchical joins run whenever their corresponding lookup input is non-`NULL`.
`spectrum_identity` receives one additional deterministic cleanup: recognizable
paths are reduced to their basename and trailing file extensions supported by
`read_any()` are removed. Numeric OPUS extensions include `.10` and any other
terminal period followed only by digits. Exact lookup keys receive the same
cleanup. Each built library records changed values and counts in its
`spectrum_identity_cleanup_report` attribute. Keep flexible class patterns in a
separate table and apply them afterward with `predict_class_reference()`.
Automatic ordinary lookups use the single shared column with overlapping values
and unique lookup keys. Lookups with no usable shared key are skipped with a
message; lookups with multiple usable shared keys are treated as ambiguous, so
wrap a lookup as `list(lookup = table, by = "key")` when an explicit key is
needed. Named `by` vectors map metadata names to different lookup names.
Canonical source metadata is standardized before these external joins: a
populated `organization` remains authoritative and a blank organization is
filled from the reviewed internal `user_name` alias. The older lookup-level
`fallback_by` field is deprecated. Set `fill_only = TRUE` to fill blank values
such as `library_type` or `spectrum_type` without replacing populated source
metadata.

```{r mini-library-build}
mini_libs <- build_lib(
  mini_files,
  recipes = list(
    raw = list(),
    derivative = list(
      conform_spec = FALSE,
      smooth_intens = TRUE,
      smooth_intens_args = list(window = 15, derivative = 1),
      make_rel = TRUE
    ),
    nobaseline = list(
      conform_spec = FALSE,
      smooth_intens = FALSE,
      subtr_baseline = TRUE,
      make_rel = TRUE
    )
  ),
  metadata_lookups = lookup,
  material_hierarchy = hierarchy,
  clean_metadata_values = TRUE,
  convert_intensity = FALSE,
  assess = TRUE,
  dedupe = FALSE
)

names(mini_libs)
check_OpenSpecy(mini_libs$raw)
check_OpenSpecy(mini_libs$derivative)
attr(mini_libs$derivative, "derivative_order")
attr(mini_libs$nobaseline, "baseline")
mini_libs$raw$metadata[
  , .(file_name, material, material_class, material_type, sn,
      assessment_flag, assessment_checks)
]
```

`prune_lib()` performs the spectrum-supported cleanup used before medoid or
model creation. It reassigns only the generic `other plastic` and
`other material` classes when an eligible same-technique candidate exists,
then evaluates material classes from largest to smallest using bounded
correlation blocks. FTIR and NIR share a candidate pool; Raman is separate;
2200--2420 cm^-1^ is excluded by default. Classes are never reduced below
`min_n`, constant or unmatched spectra are retained, and deterministic IDs
resolve correlation ties. Use `return = "report"` for the retained IDs, frozen
class schedule, reassignment events, and removals. `build_lib(prune = ...)`
maps recipe names to `prune_lib()` argument lists, so each selected spectral
representation is pruned independently.

```{r mini-library-prune}
pruned <- prune_lib(
  mini_libs$derivative,
  min_n = 1,
  return = "report",
  progress = FALSE
)
pruned$summary
pruned$schedule
```

## Official Reference-Library Workflow

The version-controlled
[`workflows/OpenSpecy_reference_library.R`](https://github.com/wincowgerDEV/OpenSpecy-package/blob/main/workflows/OpenSpecy_reference_library.R)
script now contains one default `build_lib()` call. The complete visual flow,
including checkpoint reuse and old/new assessments, is maintained in
[`.specify/memory/build-lib-diagram.html`](https://github.com/wincowgerDEV/OpenSpecy-package/blob/main/.specify/memory/build-lib-diagram.html).
Canonically named
exact-class, regex-class, library-type, material-hierarchy, and known-bad-ID
tables live under
`workflows/data/`. Raw-data corrections are completed externally before this
workflow runs. Path arguments default from `OPENSPECY_LIBRARY_DATA`,
`OPENSPECY_SOURCE_FILE`, `OPENSPECY_PROCESSED_DIR`, and
`OPENSPECY_LIBRARY_OUTPUT`, so they do not need to be assigned before the call.
The official build fills a missing organization from `user_name` before one
source-type lookup and asserts complete library/spectrum-type coverage. The
exact class lookup runs first;
`predict_class_reference()` then evaluates the separate regex table only for
blank materials. Exact/regex overlaps are audited without overwriting exact
values, and conflicting regex predictions stop for review. Remaining uncertain
identities are retained as `unclassified`. Derivative and nobaseline libraries
are pruned before their medoids and models are created; raw remains unpruned.
Each completed library, medoid, model, and assessment component is written with
a manifest under `output_dir/checkpoints`. `reuse = TRUE` loads it only when
the source/lookup signatures, arguments, package version, and builder contract
still match. Validated output is promoted into a versioned release directory,
which contains the seven legacy RDS names plus one aggregate build RDS.

The returned object has four top-level elements: `libraries`, `medoids`,
`models`, and `assessments`. Each model has one tidy `tests` table. Full
assessment uses every candidate and downloaded legacy artifact, makes one
stable-identity-grouped ten-percent holdout, removes held-out reference spectra
before identification, evaluates corresponding models, and reports clear
old/new `assess_spec(report = "all")` shifts. The smaller 1,000-spectrum run in
the manual benchmark is development-only and is not part of `build_lib()` or
final acceptance evidence.

The script is excluded from package builds but remains available
in the GitHub repository so library releases can be reviewed and reproduced.
