---
title: "Build a Mini Reference Library"
author: "Win Cowger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a Mini Reference Library}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)

data.table::setDTthreads(2)
```

```{r setup}
library(OpenSpecy)
```

This example combines a few files bundled with OpenSpecy into a small reference
library. Real library builds usually need larger lookup tables and more
curation, but the same helper functions apply.

## Read And Combine Spectra

`read_any()` can read a vector of files and pass them to `c_spec()` when the
spectra should be combined. Here the spectra are conformed to the common
wavenumber range so they can share one `OpenSpecy` object.

```{r mini-library-read}
mini_files <- c(
  read_extdata("raman_hdpe.csv"),
  read_extdata("ftir_ldpe_soil.asp"),
  read_extdata("raman_atacamit.spc")
)

mini_raw <- read_any(
  mini_files,
  c_spec_args = list(range = "common", res = 10)
)

check_OpenSpecy(mini_raw)
dim(mini_raw$spectra)
mini_raw$metadata[, "file_name", with = FALSE]
```

## Create And Fill A Lookup

Lookup templates help users see which metadata values need curation. When
`path` is not supplied, the template is returned as a `data.table`; users can
write it to CSV by supplying `path`. Lookup joins use exact values, so edit the
template until the key column matches the object metadata.

```{r mini-library-template}
template <- make_lib_lookup_template(
  mini_raw,
  columns = "file_name",
  add = c("material", "material_type")
)

template
```

For this small example, create the lookup table directly in R. The same shape
could come from a CSV edited outside R.

```{r mini-library-join}
lookup <- data.table::data.table(
  file_name = basename(mini_files),
  material = c("hdpe", "ldpe in soil", "atacamite"),
  material_type = c("plastic", "plastic", "mineral")
)

mini_raw <- join_lib_metadata(
  mini_raw,
  lookup,
  by = "file_name",
  require_complete = TRUE
)

mini_raw$metadata[, c("file_name", "material", "material_type"), with = FALSE]
```

## Build A Mini Library

`build_lib()` applies named recipes and returns a named list of libraries. This
example keeps the raw library and creates a relative-intensity version.

```{r mini-library-build}
mini_libs <- build_lib(
  mini_raw,
  recipes = list(
    raw = list(),
    relative = function(x) make_rel(x, na.rm = TRUE)
  ),
  dedupe = FALSE
)

names(mini_libs)
check_OpenSpecy(mini_libs$raw)
check_OpenSpecy(mini_libs$relative)
assess_lib(mini_libs$raw, class_col = "material_type", nearest = FALSE)
```
