---
title: "Build a Mini Reference Library"
author: "Win Cowger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a Mini Reference Library}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)

data.table::setDTthreads(2)
```

```{r setup}
library(OpenSpecy)
```

This example combines a few files bundled with OpenSpecy into a small reference
library. Real library builds usually need larger lookup tables and more
curation, but the same helper functions apply.

## Read And Combine Spectra

`c_spec()` and `build_lib()` default to the widest range represented by the
source spectra at resolution 6. Values outside each source's original range are
kept as `NA`, so useful spectral regions are not discarded.

```{r mini-library-read}
mini_files <- c(
  read_extdata("raman_hdpe.csv"),
  read_extdata("ftir_ldpe_soil.asp"),
  read_extdata("raman_atacamit.spc")
)

mini_sources <- lapply(mini_files, read_any)
mini_raw <- c_spec(mini_sources)

check_OpenSpecy(mini_raw)
dim(mini_raw$spectra)
mini_raw$metadata[, "file_name", with = FALSE]
```

## Create And Fill A Lookup

Lookup templates help users see which metadata values need curation. When
`path` is not supplied, the template is returned as a `data.table`; users can
write it to CSV by supplying `path`. Lookup joins use exact values, so edit the
template until the key column matches the object metadata.

```{r mini-library-template}
template <- make_lib_lookup_template(
  mini_raw,
  columns = "file_name",
  add = c("material", "material_type")
)

template
```

For this small example, create the lookup table directly in R. The same shape
could come from a CSV edited outside R.

```{r mini-library-join}
lookup <- data.table::data.table(
  file_name = basename(mini_files),
  library_type = c("example", "example", "example"),
  material = c("hdpe", "ldpe in soil", "atacamite")
)

hierarchy <- data.table::data.table(
  material = c("hdpe", "ldpe in soil", "atacamite"),
  material_class = c("polyethylene", "polyethylene", "copper mineral"),
  material_type = c("plastic", "plastic", "mineral")
)

join_lib_metadata(mini_raw, lookup, by = "file_name",
                  require_complete = TRUE)$metadata[
  , c("file_name", "library_type", "material"), with = FALSE
]
```

## Build A Mini Library

`build_lib()` can run the ordinary lookup and material hierarchy joins before
applying named recipes. Recipe names become names in the returned list. Empty
recipes keep the merged spectra unchanged; other recipe lists are passed to
`process_spec()`. Missing values and processing attributes are handled
automatically. Metadata column names are also cleaned to lowercase underscore
names. Known aliases are coalesced using an editable lookup table. Variants
that differ only by underscores or one terminal plural `s` match
automatically.

```{r metadata-name-lookup}
name_lookup <- lib_metadata_name_lookup(
  project_code = c("campaign id", "study code"),
  regex = list(instrument_mode = "^method_[0-9]+$")
)
name_lookup[
  canonical_name %in% c("material_color", "number_of_accumulations")
]
lib_clean_name(c("User Name", "Laser (%)", "Method...3"))
```

Named arguments add exact aliases to the defaults, while `regex` adds patterns
evaluated against cleaned names. Overlapping regex patterns produce an error
that identifies the source column and matching rules. Pass the result as
`metadata_name_lookup`. Ordinary and hierarchical joins run whenever their
corresponding lookup input is non-`NULL`.

```{r mini-library-build}
mini_libs <- build_lib(
  mini_sources,
  recipes = list(
    raw = list(),
    derivative = list(
      conform_spec = FALSE,
      smooth_intens = TRUE,
      smooth_intens_args = list(window = 15, derivative = 1),
      make_rel = TRUE
    ),
    nobaseline = list(
      conform_spec = FALSE,
      smooth_intens = FALSE,
      subtr_baseline = TRUE,
      make_rel = TRUE
    )
  ),
  metadata_lookups = lookup,
  material_hierarchy = hierarchy,
  assess = TRUE,
  dedupe = FALSE
)

names(mini_libs)
check_OpenSpecy(mini_libs$raw)
check_OpenSpecy(mini_libs$derivative)
attr(mini_libs$derivative, "derivative_order")
attr(mini_libs$nobaseline, "baseline")
mini_libs$raw$metadata[
  , .(file_name, material, material_class, material_type, sn,
      assessment_flag, assessment_checks)
]
```
