---
title: "Build a Mini Reference Library"
author: "Win Cowger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build a Mini Reference Library}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)

data.table::setDTthreads(2)
```

```{r setup}
library(OpenSpecy)
```

This example combines a few files bundled with OpenSpecy into a small reference
library. Real library builds usually need larger lookup tables and more
curation, but the same helper functions apply.

## Read And Combine Spectra

`c_spec()` and `build_lib()` default to the widest range represented by the
source spectra at resolution 6. Values outside each source's original range are
kept as `NA`, so useful spectral regions are not discarded.
This compact vignette example uses the shared overlapping range to keep the
rendered output small.

```{r mini-library-read}
mini_files <- c(
  read_extdata("raman_hdpe.csv"),
  read_extdata("ftir_ldpe_soil.asp"),
  read_extdata("raman_atacamit.spc")
)

mini_sources <- lapply(mini_files, read_any)
mini_sources <- lapply(mini_sources, function(x) {
  x$metadata$intensity_units <- "absorbance"
  attr(x, "intensity_unit") <- "absorbance"
  x
})
mini_raw <- c_spec(mini_sources, range = "common", res = 6)

check_OpenSpecy(mini_raw)
dim(mini_raw$spectra)
mini_raw$metadata[, "file_name", with = FALSE]
```

## Create And Fill A Lookup

Lookup templates help users see which metadata values need curation. When
`path` is not supplied, the template is returned as a `data.table`; users can
write it to CSV by supplying `path`. Lookup joins use exact values, so edit the
template until the key column matches the object metadata.

```{r mini-library-template}
template <- make_lib_lookup_template(
  mini_raw,
  columns = "file_name",
  add = c("material", "material_type")
)

template
```

For this small example, create the lookup table directly in R. The same shape
could come from a CSV edited outside R.

```{r mini-library-join}
lookup <- data.table::data.table(
  file_name = basename(mini_files),
  library_type = c("example", "example", "example"),
  spectrum_type = c("raman", "ftir", "raman"),
  material = c("hdpe", "ldpe in soil", "atacamite")
)

hierarchy <- data.table::data.table(
  material = c("hdpe", "ldpe in soil", "atacamite"),
  material_class = c("polyethylene", "polyethylene", "copper mineral"),
  material_type = c("plastic", "plastic", "mineral")
)

join_lib_metadata(mini_raw, lookup, by = "file_name",
                  require_complete = TRUE)$metadata[
  , c("file_name", "library_type", "spectrum_type", "material"),
  with = FALSE
]
```

## Build A Mini Library

`build_lib()` can run the ordinary lookup and material hierarchy joins before
applying named recipes. Recipe names become names in the returned list. Empty
recipes keep the merged spectra unchanged; other recipe lists are passed to
`process_spec()`. Missing values and processing attributes are handled
automatically. Metadata column names are also cleaned to lowercase underscore
names. Known aliases are coalesced using an editable lookup table. Variants
that differ only by underscores or one terminal plural `s` match
automatically.

Before sources are merged, `build_lib()` also converts declared reflectance and
transmittance spectra to absorbance by default. A nonempty
`attr(x, "intensity_unit")` is the primary truth for the whole object;
otherwise, `metadata$intensity_units` is evaluated spectrum by spectrum.
Unknown or missing units are left unchanged with a warning. Use
`convert_intensity = FALSE` when unit handling has already been completed
outside the builder.

The normal input is one or more file paths; one `OpenSpecy` or a list of
`OpenSpecy` objects supports sources already loaded in memory. An RDS path may
store either one object or a list of objects. Other formats are read with
`read_any()`. Large same-axis source lists are prepared in bulk, which is much
faster for legacy RDS files that store many one-spectrum objects. Named stages
and elapsed time are reported while the library is built; use
`progress = FALSE` when quiet output is preferred. Supplying
`restrict_range_args` triggers the existing `restrict_range()` operation before
deduplication and recipes; multiple retained ranges can exclude a known silent
region without custom workflow code.

```{r metadata-name-lookup}
name_lookup <- lib_metadata_name_lookup(
  project_code = c("campaign id", "study code"),
  regex = list(instrument_mode = "^method_[0-9]+$")
)
name_lookup[
  canonical_name %in% c("material_color", "number_of_accumulations")
]
lib_clean_name(c("User Name", "Laser (%)", "Method...3"))
```

Named arguments add exact aliases to the defaults, while `regex` adds patterns
evaluated against cleaned names. Overlapping regex patterns produce an error
that identifies the source column and matching rules. Pass the result as
`metadata_name_lookup`. Set `clean_metadata_values = TRUE` in `build_lib()`
or `clean_values = TRUE` in `lib_clean_metadata()` to lowercase, trim, and
ASCII-normalize character metadata values before joins. Ordinary and
hierarchical joins run whenever their corresponding lookup input is non-`NULL`.
`spectrum_identity` receives one additional deterministic cleanup: recognizable
paths are reduced to their basename and trailing file extensions supported by
`read_any()` are removed. Numeric OPUS extensions include `.10` and any other
terminal period followed only by digits. Exact lookup keys receive the same
cleanup. Each built library records changed values and counts in its
`spectrum_identity_cleanup_report` attribute. Keep flexible class patterns in a
separate table and apply them afterward with `predict_class_reference()`.
Automatic ordinary lookups use the single shared column with overlapping values
and unique lookup keys. Lookups with no usable shared key are skipped with a
message; lookups with multiple usable shared keys are treated as ambiguous, so
wrap a lookup as `list(lookup = table, by = "key")` when an explicit key is
needed. Named `by` vectors map metadata names to different lookup names.
Canonical source metadata is standardized before these external joins: a
populated `organization` remains authoritative and a blank organization is
filled from the reviewed internal `user_name` alias. The older lookup-level
`fallback_by` field is deprecated. Set `fill_only = TRUE` to fill blank values
such as `library_type` or `spectrum_type` without replacing populated source
metadata.

```{r mini-library-build}
mini_libs <- build_lib(
  mini_files,
  recipes = list(
    raw = list(),
    derivative = list(
      conform_spec = FALSE,
      smooth_intens = TRUE,
      smooth_intens_args = list(window = 15, derivative = 1),
      make_rel = TRUE
    ),
    nobaseline = list(
      conform_spec = FALSE,
      smooth_intens = FALSE,
      subtr_baseline = TRUE,
      make_rel = TRUE
    )
  ),
  metadata_lookups = lookup,
  material_hierarchy = hierarchy,
  clean_metadata_values = TRUE,
  convert_intensity = FALSE,
  assess = TRUE,
  dedupe = FALSE
)

names(mini_libs)
check_OpenSpecy(mini_libs$raw)
check_OpenSpecy(mini_libs$derivative)
attr(mini_libs$derivative, "derivative_order")
attr(mini_libs$nobaseline, "baseline")
mini_libs$raw$metadata[
  , .(file_name, material, material_class, material_type, sn,
      assessment_flag, assessment_checks)
]
```

`prune_lib()` performs the spectrum-supported cleanup used before medoid or
model creation. Within the same spectral-technique pool, `other` may be
reassigned to the nearest established class, `other plastic` only to a
candidate whose `material_type` is `plastic`, and `other material` only to
`organic matter` or `mineral`. The matched `material_type` is copied with the
class. It then evaluates material classes from largest to smallest using
bounded correlation blocks. FTIR and NIR share a candidate pool; Raman is
separate; 2200--2420 cm^-1^ is excluded by default. After generic reassignment,
each `spectrum_type` and material-class group must contain at least `min_n`
spectra. Smaller groups are removed before correlation pruning, while groups
exactly at the threshold are retained whole and larger groups are never reduced
below it. Constant or unmatched spectra in eligible classes are retained, and
deterministic IDs resolve correlation ties. Use `return = "report"` for the
retained IDs, frozen class schedule, reassignment correlations, spectrum-level
removals, and `excluded_classes`, which reports observed support and the number
of additional or reassigned spectra needed. End-to-end builds combine these
class rows in `assessments$pruning_excluded_classes` with the recipe name.
`build_lib(prune = ...)` maps recipe names to `prune_lib()` argument lists, so
each selected spectral representation is pruned independently. In the official
workflow, an otherwise unresolved identity is temporarily labeled `other`.
The default `remove_other = TRUE` removes blank identities and the generic
`other`, `other plastic`, and `other material` labels before quality control,
while keeping their identifiers, source metadata, prior labels, reasons, and
actions in `assessments$other_review`; `assessments$other_filter` holds typed
before/after counts. Set `remove_other = FALSE` to retain these rows for the
constrained semisupervised `prune_lib()` reassignment described above.

```{r mini-library-prune}
pruned <- prune_lib(
  mini_libs$derivative,
  min_n = 1,
  return = "report",
  progress = FALSE
)
pruned$summary
pruned$schedule
pruned$excluded_classes
```

## Official Reference-Library Workflow

The version-controlled
[`workflows/OpenSpecy_reference_library.R`](https://github.com/wincowgerDEV/OpenSpecy-package/blob/main/workflows/OpenSpecy_reference_library.R)
script retains the source and output path finders and passes those paths to one
`build_lib()` call. The complete visual flow,
including checkpoint reuse and old/new assessments, is maintained in
[`.specify/memory/build-lib-diagram.html`](https://github.com/wincowgerDEV/OpenSpecy-package/blob/main/.specify/memory/build-lib-diagram.html).
Canonically named
exact-class, regex-class, library-type, material-hierarchy, and known-bad-ID
tables live under
`workflows/data/`. Raw-data corrections are completed externally before this
workflow runs. The source-library paths and output directory are explicit
inputs. When `workflow_data` is not supplied, `build_lib()` looks for the
helper tables under `data/` beside the calling script, then under `data/` or
`workflows/data/` in the current working directory. Calling `build_lib()` with
no `x` now stops with an actionable error instead of guessing external paths.
The official build fills a missing organization from `user_name` before one
source-type lookup and asserts complete library/spectrum-type coverage. The
exact class lookup runs first;
`predict_class_reference()` then evaluates the separate regex table only for
blank materials. Exact/regex overlaps are audited without overwriting exact
values, and conflicting regex predictions stop for review. Material classes
use concise polymer-family names, with polyethylene and polypropylene separated
and `polyhydroxy(meth)acrylates` retaining its chemically meaningful optional
group notation. Remaining uncertain identities follow the selected generic-row
review policy. Derivative and
nobaseline spectra pass three quality gates before pruning: FTIR CO2 is
flattened when the 2200--2420 maximum divided by the 2420--2550 silent-region
maximum is greater than two; high-tail detection ignores NA padding, trims each
spectrum on its finite support, and drops failed corrections; and finite
running SNR below two (or unavailable SNR) is removed. Raw remains unpruned.
Each completed library, medoid, model, and assessment component is written with
a manifest under `output_dir/checkpoints`. `reuse = TRUE` loads it only when
the source/lookup signatures, arguments, package version, and builder contract
still match. Validated output is promoted into a versioned release directory,
which contains the seven legacy RDS names, explicit algorithm-named model RDS
files, and one aggregate build RDS.

The returned object has four top-level elements: `libraries`, `medoids`,
`models`, and `assessments`. Libraries are recipe-first and then keyed by
`ftir`, `raman`, or `nir`: full Raman uses 200--4000, FTIR uses 400--4000,
and NIR uses 4000--12000 cm^-1^. FTIR/Raman medoids and models use 800--3200;
the NIR identification interval is derived from the longest region with at
least 90% finite coverage (falling back to available coverage) and recorded on
the object. All-blank metadata columns are dropped independently by type.
Each spectrum must observe at least 10% of its type-specific identification
axis to enter medoid selection or model fitting. PAM operates on a temporary
relative matrix where each spectrum's missing values are filled by that
spectrum's finite mean. Selected medoid IDs are then pulled from the original
range-restricted object, so published medoids retain their original `NA`
positions. `models` is algorithm-first (`logistic_regression` or
`random_forest`), then recipe and type. Logistic regression retains
derivative/nobaseline medoid training and fills restored gaps with the finite
mean at each wavenumber across training medoids. Random forests use the complete
eligible raw, derivative, and nobaseline type libraries, relative normalization,
wavenumber-mean filling, inverse-frequency balanced case sampling, 500
probability trees, permutation importance, and all available ranger threads.
Balanced sampling improves minority representation in each bootstrap without
also applying a class-vote correction. Because this resampling can make
out-of-bag accuracy optimistic, treat OOB results as fit diagnostics and use
the grouped held-out assessment as the model comparison and selection result.
`train_spec_model()` exposes both engines; `build_model_lib()` remains a
compatible wrapper. Each model stores its filler, support audit, training
diagnostics, and one tidy `tests` table. Logistic lambda selection remains
maximum out-of-fold macro class accuracy with its calibrated alpha,
no-intercept, grouped multinomial, and weighting settings unchanged.

Full assessment uses every candidate and downloaded legacy artifact. Candidate
artifacts use a seeded approximately ten-percent class/type-stratified sample
from the candidate full library, while legacy artifacts independently use the
legacy full library. This source-local policy tolerates taxonomy evolution
without fuzzy cross-version class matching; its results describe each
artifact on its own source population and are not paired causal deltas.
Held-out identifiers are removed from full and medoid references before
identification. Candidate random forests are assessed by fitting a separate
forest on only the 90% full-library training partition and predicting its held
out 10%; their production forests still use all eligible spectra. Logistic and
published artifacts retain their existing source-local evaluation contract.
All model queries come from full libraries, never medoids, and `match_spec()`
uses stored fillers for partial spectra. Macro class
accuracy is the primary metric, followed by evaluated coverage, overall
accuracy, per-class accuracy, and clear old/new
`assess_spec(report = "all")` shifts. Empty warnings retain their column schema
and numeric measurements remain numeric. Confusion tables mark incorrect cells,
report each cell's fraction of its expected class, and order the largest
misidentifications first. The smaller 1,000-spectrum run in the
manual benchmark is development-only and is not part of `build_lib()` or final
acceptance evidence.

The standard call therefore has this shape:

```{r official-call-shape, eval=FALSE}
reference_library_build <- build_lib(
  files,
  output_dir = output_dir,
  previous_library_dir = "system",
  remove_other = TRUE
)
```

After upstream libraries have already completed, rebuild only the downstream
artifacts into a new explicit output location:

```{r downstream-rebuild-shape, eval=FALSE}
updated_reference_library_build <- rebuild_lib_artifacts(
  completed_build_dir,
  output_dir = downstream_output_dir,
  previous_library_dir = "system"
)
```

`completed_build_dir` may instead be a `reference_library_build.rds`, a
`libraries.rds` checkpoint, or the corresponding in-memory build object.

The script is excluded from package builds but remains available
in the GitHub repository so library releases can be reviewed and reproduced.
