---
title: "Masked Circular Autoencoder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Masked Circular Autoencoder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("torch", quietly = TRUE)
)
```

The masked circular autoencoder is an experimental workflow for compressing
OpenSpecy-style spectra into one circular coordinate. The model uses spectral
values and observed-value masks only. Polymer names, classes, instruments, or
other metadata are reserved for post-training diagnostics and plots.

```{r setup}
library(OpenSpecy)
```

The package includes `test_lib`, a compact reference library suitable for small
examples. Larger manual experiments can use `get_lib("derivative")` followed by
`load_lib("derivative")` outside routine test runs.

```{r data}
data("test_lib")
os <- test_lib
os$spectra <- os$spectra[, seq_len(12)]
os$metadata <- os$metadata[seq_len(12), ]
```

Fit a small masked circular model. Missing values are treated as unobserved
spectral positions, not as zero intensity.

```{r fit}
model <- fit_masked_circular_ae(
  os,
  n_hidden = c(32, 8),
  min_overlap_points = 20,
  min_overlap_fraction = 0.1,
  epochs = 5,
  batch_size = 4,
  verbose = FALSE,
  seed = 1
)
```

Encode spectra, reconstruct them on the model grid, and inspect diagnostics.

```{r encode}
encoded <- encode_masked_circular_ae(model, os)
reconstructed <- reconstruct_masked_circular_ae(model, x = os)
diagnostics <- diagnose_masked_circular_ae(model, os, encoded = encoded,
                                           n_pairs = 100, k = 3)
head(encoded)
diagnostics$angular_spread
```

Metadata can be used after fitting for diagnostics or display.

```{r plot}
plot_circular_embedding(encoded, color_by = "sample_name",
                        metadata = os$metadata)
```
